---
title: "Comment on 'Age of enlightenment: long-term effects of outdoor aesthetic lights on bats in churches'"
site: bookdown::bookdown_site
output: 
  bookdown::pdf_book:
    base_format: INBOmd::rsos_article

author:
  - name: T Onkelinx
    affiliation: "1"
address:
  - code: "1"
    address: Research Institute for Nature and Forest, Kliniekstraat 25, 1070 Anderlecht, Belgium

corresp_author_name:  "T. Onkelinx"
corresp_author_email: "thierry.onkelinx@inbo.be"

subject:
  - "ecology"
  - "environmental science"
  - "behaviour"

keywords:
  - global change
  - energy
  - cultural heritage
  - historic buildings
  - biodiversity
  - light pollution

abstract: |
  This comment reanalyses the data presented in [@Rydell] which wre analyses using only very basic statistics as Fisher's extact test and McNemar's test. We would like to demonstrate how the use of more advanced statistical methods can make better use of the available data, quantify the observed effects and strengthen the conclusions in [@Rydell].

first_page_text: |

## Remove this if not required
data_accessibility: |
  The data and the code is available at https://github.com/thierryo/rydell_etal_2017

author_contributions: |
  T.O. made the analysis and wrote the draft.

## Remove this if not required
conflict_of_interest: |
  The authors have no competing interests.

bibliography: bibliography.bib

## change to true to add optional line numbering
lineno: false

---

```{r load-packages, include = FALSE}
library(knitr)
opts_chunk$set(
  cache = TRUE,
  autodep = TRUE,
  fig.width = 4,
  fig.height = 2,
  echo = FALSE,
  message = FALSE
)
library(tidyverse)
library(ordinal)
library(scales)
theme_set(theme_grey(base_size = 6))
```

```{r load-data}
status <- read_csv("../../data/status.csv") %>%
  mutate(
    Church = factor(Church),
    Lights = factor(
      Lights,
      levels = c("not", "half", "full"),
      labels = c("dark", "partly lit", "fully lit")
    ),
    Period = factor(
      Period,
      levels = c("80's", "2016")
    ),
    Status = factor(
      Status,
      levels = c("colony", "used", "no bats"),
      ordered = TRUE
    )
  )
```

```{r}
contrasts <- function(model, contrast, n = 1000){
  cf <- summary(model)$coef
  params <- rnorm(
    n * nrow(cf), 
    mean = cf[, "Estimate"], 
    sd = cf[, "Std. Error"]
  ) %>%
    matrix(nrow = nrow(cf))
  contrast %*% params
}
```

# The 2016 survey

```{r current-fisher}
current_cont <- status %>%
  filter(Period == "2016")
current_p <- fisher.test(current_cont$Status, current_cont$Lights)$p.value %>%
  round(4)
```

The 2016 survey hold information on the presence of bats and the lights on the churches. Both variables can be represented as ordinal variables. The presence of the bats is recorded as `no bats` < `used` < `colony`. The light status is `dark` < `partly lit` < `fully lit`. After calculating a contigency table (tab \@ref(tab:current-contigency)), we can apply Fisher's exact test ($p = `r current_p`$). The interpretation of the value is limited: we reject the null hypothesis that the presence of bats and the light status are independent. It doesn't yield information on the effect size. To get arround this, the 3x3 contingency table was recombined in several 2x2 tables. Each of those were subjected to Fisher's extact test and the effect size was based on interpretation of the 2x2 tables. We found 2 such test in the manuscript [@Rydell] and 4 in the dataset.

```{r current-contigency}
crosstab <- current_cont %>%
  mutate(
    Lights = as.character(Lights),
    Status = as.character(Status)
  ) %>%
  count(Lights, Status)
crosstab %>%
  group_by(Lights) %>%
  summarise(
    Status = "total",
    n = sum(n)
  ) %>%
  bind_rows(
    crosstab %>%
      group_by(Status) %>%
      summarise(
        Lights = "total",
        n = sum(n)
      ),
    crosstab %>%
      summarise(
        Status = "total",
        Lights = "total",
        n = sum(n)
      ),
    crosstab
  ) %>%
  mutate(
    Status = factor(
      Status, 
      levels = c("no bats", "used", "colony", "total")
    ),
    Lights = factor(
      Lights, 
      levels = c("dark", "partly lit", "fully lit", "total")
    )
  ) %>%
  spread(Status, n) %>%
  kable(caption = "Contingenty table for the 2016 survey")
```

```{r current-ordinal}
current_model <- clm(Status ~ Lights, data = status, subset = Period == "2016")
current_lrt <- drop1(current_model, test = "Chisq")
current_lrt <- sprintf(
  "$\\chi^2_%i$=%.4g, p = %.2g", 
  current_lrt[2, "Df"], 
  current_lrt[2, "LRT"], 
  current_lrt[2, "Pr(>Chi)"]
)
```

Ordinal data can be analysed with cumulative logit models[@ordinal]. The general form is given in eq \@ref(eq:cum-logit). In this case the presence of bats is the response variable and the light status is used as covariate. We can use a likelihood ratio test to test for the overall effect of the light status (`r current_lrt`). The effect size is quantified through the coefficients (tab. \@ref(tab:current-coef)). The predictions (fig. \@ref(fig:current-prediction)) yield the same values as fig. 2 in [@Rydell] but fig. \@ref(fig:current-prediction) adds confidence intervals on top.

\begin{equation} 
  \mbox{logit}(P(Y_i <= j)) = \theta_j - x_i^T\beta
  (\#eq:cum-logit)
\end{equation} 

```{r current-coef}
current_or <- confint(current_model) %>%
  cbind(mean = coef(current_model)[3:4]) %>%
  exp() %>%
  as.data.frame() %>%
  select(mean = 3, lcl = 1, ucl = 2) %>%
  mutate(
    print = sprintf("$%.2f (%.2f; %.2f)$", mean, lcl, ucl)
  )
current_model %>%
  summary() %>%
  coef() %>%
  kable(
    digits = c(3, 3, 2, 4),
    caption = "Coefficients of the cumulative logit model for the 2016 survey."
  )
```

## Interpretation

Placing lights all around the exterior has strong negative effects on the presences of bats. The odds ratio for having no bats when fully lit is `r current_or$print[2]` compared to dark churches. The difference between dark and partly lit churches is unclear (odds ratio: `r current_or$print[1]`).

```{r current-contrast}
cumulative_logit <- contrasts(
  current_model, 
  contrast =  rbind(
    "colony:dark" = c(1, 0, 0, 0),
    "colony:partly lit" = c(1, 0, -1, 0),
    "colony:fully lit" = c(1, 0, 0, -1),
    "colony or used:dark" = c(0, 1, 0, 0),
    "colony or used:partly lit" = c(0, 1, -1, 0),
    "colony or used:fully lit" = c(0, 1, 0, -1)
  )
)
cumulative_prediction <- cumulative_logit %>%
  plogis() %>%
  apply(
    1, 
    function(x){
      c(
        Mean = mean(x), 
        LCL = quantile(x, 0.025) %>%
          unname(), 
        UCL = quantile(x, 0.975) %>%
          unname()
      )
    }
  ) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("Contrast") %>%
  extract(Contrast, c("Status", "Lights"), "(.*):(.*)") %>%
  mutate(Lights = factor(Lights, levels = levels(status$Lights)))
```

```{r current-prediction, fig.cap = "Estimated cumulative probability based on the 2016 survey."}
ggplot(
  cumulative_prediction, 
  aes(
    x = Lights, 
    y = Mean, ymin = LCL, ymax = UCL, 
    colour = Status
  )
) +
  geom_errorbar(position = position_dodge(width = 0.6)) +
  geom_point(position = position_dodge(width = 0.6)) +
  scale_y_continuous(
    "Estimated cumulative probability",
    labels = percent, 
    limits = 0:1
  ) +
  scale_color_manual(
    values = c("colony" = "black", "colony or used" = "darkgrey")
  )
```

# Comparison with 1980s survey

```{r}
status_model <- clmm(Status ~ Period + Lights + (1|Church), data = status)
drop1(status_model, test = "Chisq")
summary(status_model)
```

```{r}
cumulative_logit <- contrasts(
  status_model, 
  contrast =  rbind(
    "colony:80's:dark" = c(1, 0, 0, 0, 0),
    "colony:2016:dark" = c(1, 0, -1, 0, 0),
    "colony:2016:partly lit" = c(1, 0, -1, -1, 0),
    "colony:2016:fully lit" = c(1, 0, -1, 0, -1),
    "colony or used:80's:dark" = c(0, 1, 0, 0, 0),
    "colony or used:2016:dark" = c(0, 1, -1, 0, 0),
    "colony or used:2016:partly lit" = c(0, 1, -1, -1, 0),
    "colony or used:2016:fully lit" = c(0, 1, -1, 0, -1)
  )
)
cumulative_prediction <- cumulative_logit %>%
  plogis() %>%
  apply(
    1, 
    function(x){
      c(
        Mean = mean(x), 
        LCL = quantile(x, 0.025) %>%
          unname(), 
        UCL = quantile(x, 0.975) %>%
          unname()
      )
    }
  ) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("Contrast") %>%
  extract(Contrast, c("Status", "Period", "Lights"), "(.*):(.*):(.*)")
```

```{r time-model, fig.cap = "Estimated cumulative probability"}
ggplot(
  cumulative_prediction, 
  aes(
    x = Lights, 
    y = Mean, ymin = LCL, ymax = UCL, 
    colour = Status, 
    shape = Period, linetype = Period
  )
) +
  geom_errorbar(position = position_dodge(width = 0.5)) +
  geom_point(position = position_dodge(width = 0.5)) +
  scale_y_continuous(
    "Estimated cumulative probability",
    labels = percent, 
    limits = 0:1
  ) + 
  coord_flip()
```

# Effect of renovations

```{r}
status %>%
  filter(Period == "2016") %>%
  count(Lights, Renovated, Status)
```


```{r}
renovation_model <- clmm(
  Status ~ Period + Lights + Renovated + (1|Church), 
  data = status
)
drop1(renovation_model, test = "Chisq")
summary(renovation_model)
```

```{r}
cumulative_logit <- contrasts(
  renovation_model, 
  contrast =  rbind(
    "colony:80's:dark:no" = c(1, 0, 0, 0, 0, 0),
    "colony or used:80's:dark:no" = c(0, 1, 0, 0, 0, 0),
    "colony:2016:dark:no" = c(1, 0, -1, 0, 0, 0),
    "colony:2016:partly lit:no" = c(1, 0, -1, -1, 0, 0),
    "colony:2016:fully lit:no" = c(1, 0, -1, 0, -1, 0),
    "colony or used:2016:dark:no" = c(0, 1, -1, 0, 0, 0),
    "colony or used:2016:partly lit:no" = c(0, 1, -1, -1, 0, 0),
    "colony or used:2016:fully lit:no" = c(0, 1, -1, 0, -1, 0),
    "colony:2016:dark:yes" = c(1, 0, -1, 0, 0, -1),
    "colony:2016:partly lit:yes" = c(1, 0, -1, -1, 0, -1),
    "colony:2016:fully lit:yes" = c(1, 0, -1, 0, -1, -1),
    "colony or used:2016:dark:yes" = c(0, 1, -1, 0, 0, -1),
    "colony or used:2016:partly lit:yes" = c(0, 1, -1, -1, 0, -1),
    "colony or used:2016:fully lit:yes" = c(0, 1, -1, 0, -1, -1)
  )
)
cumulative_prediction <- cumulative_logit %>%
  plogis() %>%
  apply(
    1, 
    function(x){
      c(
        Mean = mean(x), 
        LCL = quantile(x, 0.025) %>%
          unname(), 
        UCL = quantile(x, 0.975) %>%
          unname()
      )
    }
  ) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("Contrast") %>%
  extract(
    Contrast, 
    c("Status", "Period", "Lights", "Renovation"), 
    "(.*):(.*):(.*):(.*)"
  ) %>%
  mutate(
    Period = ifelse(
      Period == "2016",
      Renovation,
      Period
    ) %>%
      factor(
        levels = c("80's", "no", "yes"),
        labels = c("80's", "2016", "2016 with\nrenovation")
      ),
    Lights = factor(Lights, levels = levels(status$Lights))
  )
```

```{r renovation-model, fig.cap = "Estimated cumulative probability"}
ggplot(
  cumulative_prediction, 
  aes(
    x = Lights, 
    y = Mean, ymin = LCL, ymax = UCL, 
    colour = Period
  )
) +
  geom_errorbar(position = position_dodge(width = 0.5)) +
  geom_point(position = position_dodge(width = 0.5)) +
  scale_x_discrete("Light status 2016") +
  scale_y_continuous(
    "Estimated cumulative probability",
    labels = percent, 
    limits = 0:1
  ) + 
  facet_wrap(~Status)
```

# Conclusions
