---
title: "Comment on 'Age of enlightenment: long-term effects of outdoor aesthetic lights on bats in churches'"
site: bookdown::bookdown_site
output: 
  bookdown::pdf_book:
    base_format: rticles::rsos_article

author:
  - name: Thierry Onkelinx
    affiliation: "1"
address:
  - code: "1"
    address: Research Institute for Nature and Forest, Kliniekstraat 25, 1070 Anderlecht, Belgium

corresp_author_name:  "Thierry Onkelinx"
corresp_author_email: "thierry.onkelinx@inbo.be"

subject:
  - "ecology"
  - "environmental science"
  - "behaviour"

keywords:
  - global change
  - energy
  - cultural heritage
  - historic buildings
  - biodiversity
  - light pollution
  - ordinal regression
  - mixed models

abstract: |
  This comment reanalyses the data presented in [@Rydell] which were analysed using only very basic statistics like Fisher's extact test and McNemar's test. We demonstrate how the use of more advanced statistical methods can make better use of the available data, quantify the observed effects and strengthen the conclusions in [@Rydell]. We have no intention to discredit the origal authors. Their analyses were basic but correct.

## Remove this if not required
data_accessibility: |
  The data and the code is available on https://github.com/ThierryO/rydell_etal_2017 (DOI: [10.5281/zenodo.1002852](https://doi.org/10.5281/zenodo.1002852))

author_contributions: |
  T.O. made the analysis and wrote the draft.

## Remove this if not required
conflict_of_interest: |
  The authors have no competing interests.

bibliography: bibliography.bib

## change to true to add optional line numbering
lineno: false

---

```{r load-packages, include = FALSE}
library(knitr)
opts_chunk$set(
  cache = TRUE,
  autodep = TRUE,
  fig.width = 4,
  fig.height = 1.6,
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
library(tidyverse)
library(ordinal)
library(mvtnorm)
library(scales)
theme_set(
  theme_grey(base_size = 6) +
  theme(
    legend.title = element_text(size = rel(0.8)),
    legend.spacing = unit(0.1, "cm"),
    legend.key.size = unit(1, "lines")
  )
)
```

```{r load-data}
status <- read_csv("../../data/status.csv") %>%
  mutate(
    Church = factor(Church),
    Lights = factor(
      Lights,
      levels = c("not", "half", "full"),
      labels = c("dark", "partly lit", "fully lit")
    ),
    Survey = factor(
      Survey,
      levels = c("80's", "2016"),
      labels = c("1980s", "2016")
    ),
    Renovated = factor(
      Renovated, 
      levels = c(FALSE, TRUE), 
      labels = c("no", "yes")
    ),
    Status = factor(
      Status,
      levels = c("colony", "used", "no bats"),
      ordered = TRUE
    )
  ) %>%
  add_count(Church)
```

```{r simulate_contrast}
contrasts <- function(model, contrast, n = 1000){
  model %>%
    vcov %>%
    "["(names(model$coefficients), names(model$coefficients)) %>%
    rmvnorm(n = n, mean = model$coefficients) %>%
    tcrossprod(x = contrast)
}
```
# Material and methods

The analysis was based on the original data[@Rydell] and run in R 3.4.1 [@R]. The `tidyverse`[@tidyverse] collection of packages was used for the data import, data wrangling and data visualisation. `ordinal`[@ordinal] was used for the advanced analyses.

# Results

## The 2016 survey

```{r current-fisher}
current_cont <- status %>%
  filter(Survey == "2016")
current_p <- fisher.test(current_cont$Status, current_cont$Lights)$p.value %>%
  round(4)
```

The 2016 survey holds information on the presence of bats and the light status of the churches. Both variables can be represented as ordinal variables. The presence of the bats is recorded as `no bats` < `used` < `colony`. The light status is `dark` < `partly lit` < `fully lit`. After calculating a contigency table, we can apply Fisher's exact test ($p = `r current_p`$). The interpretation of the test is limited: we reject the null hypothesis that the presence of bats and the light status are independent. It doesn't yield information on the effect size. To get arround this, the 3x3 contingency table was recombined in several 2x2 tables. Each of those were subjected to Fisher's extact test and the effect size was based on interpretation of the 2x2 tables. We found two such tests in the manuscript [@Rydell] and four in the dataset.

\EndFirstPage

```{r current-contigency}
crosstab <- current_cont %>%
  transmute(
    Lights = as.character(Lights),
    Status = as.character(Status)
  ) %>%
  count(Lights, Status)
```

```{r current-ordinal}
current_model <- clm(Status ~ Lights, data = status, subset = Survey == "2016")
current_lrt <- drop1(current_model, test = "Chisq")
current_lrt <- sprintf(
  "$\\chi^2_%i$=%.4g, p = %.2g", 
  current_lrt[2, "Df"], 
  current_lrt[2, "LRT"], 
  current_lrt[2, "Pr(>Chi)"]
)
```

Another option is to analyse ordinal data with cumulative logit models[@ordinal]. The general form is given in eq \@ref(eq:cum-logit). In this case the presence of bats is the response variable and the light status is the covariate. We test for the overall effect of the light status with a likelihood ratio test ($LRT$) (`r current_lrt`). The predictions (fig. \@ref(fig:current-prediction)) yield similar values as fig. 2 in [@Rydell].

\begin{equation} 
  \mbox{logit}(P(Y_i <= j)) = \theta_j - x_i^T\beta
  (\#eq:cum-logit)
\end{equation} 

```{r current-coef}
current_or <- confint(current_model) %>%
  cbind(mean = coef(current_model)[3:4]) %>%
  exp() %>%
  as.data.frame() %>%
  select(mean = 3, lcl = 1, ucl = 2) %>%
  mutate(
    print = sprintf("$OR:%.2f (%.2f; %.2f)$", mean, lcl, ucl)
  )
```

### Interpretation

Placing lights all around the exterior has strong negative effects on the presence of bats. The odds ratio ($OR$) for having no bats when fully lit compared to dark churches is `r current_or$print[2]`. The difference between dark and partly lit churches is unclear^[Note that $1$ is included in the confidence interval. $OR = 1$ implies that both odds are equal, hence no effect.] `r current_or$print[1]`.

```{r current-contrast}
cumulative_logit <- contrasts(
  current_model, 
  contrast = rbind(
    "colony:dark" = c(1, 0, 0, 0),
    "colony:partly lit" = c(1, 0, -1, 0),
    "colony:fully lit" = c(1, 0, 0, -1),
    "colony or used:dark" = c(0, 1, 0, 0),
    "colony or used:partly lit" = c(0, 1, -1, 0),
    "colony or used:fully lit" = c(0, 1, 0, -1)
  )
)
cumulative_prediction <- cumulative_logit %>%
  plogis() %>%
  apply(
    1, 
    function(x){
      c(
        Mean = mean(x), 
        LCL = quantile(x, 0.025) %>%
          unname(), 
        UCL = quantile(x, 0.975) %>%
          unname()
      )
    }
  ) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("Contrast") %>%
  extract(Contrast, c("Status", "Lights"), "(.*):(.*)") %>%
  mutate(Lights = factor(Lights, levels = levels(status$Lights)))
```

```{r current-prediction, fig.cap = "Estimated cumulative probability based on the 2016 survey using the basic and advanced method."}
status %>%
  filter(Survey == "2016", Status != "no bats") %>%
  count(Lights, Status) %>%
  group_by(Lights) %>%
  mutate(
    Naive = cumsum(nn),
    Status = ifelse(Status == "colony", "colony", "colony or used")
  ) %>%
  ungroup() %>%
  inner_join(
    status %>%
      filter(Survey == "2016") %>%
      count(Lights),
    by = "Lights"
  ) %>%
  transmute(
    Lights,
    Status,
    Mean = Naive / nn.y,
    Method = "basic"
  ) %>%
  bind_rows(
    cumulative_prediction %>%
      mutate(Method = "advanced")
  ) %>%
  mutate(
    Method = factor(Method, levels = c("basic", "advanced"))
  ) %>%
  ggplot(
    aes(
      x = Lights, 
      y = Mean, ymin = LCL, ymax = UCL, 
      colour = Status,
      shape = Method
    )
  ) +
  geom_errorbar(position = position_dodge(width = 0.5)) +
  geom_point(position = position_dodge(width = 0.5)) +
  scale_x_discrete("Light status") +
  scale_y_continuous(
    "Estimated cumulative probability",
    labels = percent, 
    limits = 0:1
  ) +
  scale_shape_manual(values = c("advanced" = 19, "basic" = 4)) +
  scale_color_manual(
    values = c("colony" = "black", "colony or used" = "darkgrey")
  )
```

## Comparison with 1980s survey

During the 2016 survey, all 60 churches of the 1980s survey have been revisted and 50 additional churches. The orignal authors take the repeated measures into account by using McNemar's test[@Rydell]. This test works only on 2x2 tables. 

A better alternative would be to use cumulative logit mixed models[@ordinal]. In this case we use survey and light status as fixed effects and church as a random effect. This random effect takes the paired nature of the dataset into account. Note that the model separates the survey effect and the light status effect. The survey effect would an overall decline in bat between surveys when the church remains dark. The interaction between survey and light status is not relevant since all churches were unlit in the 1980s survey. While the McNemar test requires all data to be paired, the mixed model does not. Hence we can also use the data on the churches surveyed only in 2016.

```{r survey-ordinal}
survey_model0 <- clmm(
  Status ~ Survey + Lights + (1|Church), 
  data = status, 
  subset = n == 2
)
survey_model <- clmm(Status ~ Survey + Lights + (1|Church), data = status)
survey_lrt0 <- drop1(survey_model0, test = "Chisq")
survey_lrt0 <- sprintf(
  "$\\chi^2_%i$=%.4g, p = %.2g", 
  survey_lrt0[2:3, "Df"], 
  survey_lrt0[2:3, "LRT"], 
  survey_lrt0[2:3, "Pr(>Chi)"]
)
survey_lrt <- drop1(survey_model, test = "Chisq")
survey_lrt$`Pr(>Chi)` <- ifelse(
  survey_lrt$`Pr(>Chi)` < 0.0001,
  "p < 0.0001",
  sprintf("p = %.2g", survey_lrt$`Pr(>Chi)`)
)
survey_lrt <- sprintf(
  "$\\chi^2_%i$=%.4g, %s", 
  survey_lrt[2:3, "Df"], 
  survey_lrt[2:3, "LRT"], 
  survey_lrt[2:3, "Pr(>Chi)"]
)
```

Let's see what the impact of using a larger dataset is. The $LRT$ for the survey effect on the paired data is `r survey_lrt0[1]` and `r survey_lrt[1]` for the full data. Likewise we have `r survey_lrt0[2]` and `r survey_lrt[2]` for the light status. The influence on the predictions is depicted in fig. \@ref(fig:survey-prediction). The differences in predictions might be due a possible bias in the 1980s survey, since the paired data yields higher estimates. Note that the confidence interval on the predictions are more narrow when we use the full data and that the estimates of the paired dataset are within the confidence intervals of the full data.

```{r survey-coef}
survey_or <- confint(survey_model) %>%
  cbind(mean = coef(survey_model)) %>%
  exp() %>%
  as.data.frame() %>%
  select(mean = 3, lcl = 1, ucl = 2) %>%
  mutate(
    print = sprintf("$OR:%.2f (%.2f; %.2f)$", mean, lcl, ucl)
  )
```

### Interpretation^[Based on the full data.]

The difference in odds for finding no bats between both surveys is unclear `r survey_or$print[3]`. Likewise is the difference between dark and partly lit churches in 2016 `r survey_or$print[4]`. The odds highly increase when changing for a dark to a full lit church `r survey_or$print[5]`.

```{r survey-contrast}
cumulative_logit <- contrasts(
  survey_model, 
  contrast =  rbind(
    "all:colony:1980s:dark" = c(1, 0, 0, 0, 0),
    "all:colony:2016:dark" = c(1, 0, -1, 0, 0),
    "all:colony:2016:partly lit" = c(1, 0, -1, -1, 0),
    "all:colony:2016:fully lit" = c(1, 0, -1, 0, -1),
    "all:colony or used:1980s:dark" = c(0, 1, 0, 0, 0),
    "all:colony or used:2016:dark" = c(0, 1, -1, 0, 0),
    "all:colony or used:2016:partly lit" = c(0, 1, -1, -1, 0),
    "all:colony or used:2016:fully lit" = c(0, 1, -1, 0, -1)
  )
) %>%
  rbind(
    contrasts(
      survey_model0, 
      contrast =  rbind(
        "paired:colony:1980s:dark" = c(1, 0, 0, 0, 0),
        "paired:colony:2016:dark" = c(1, 0, -1, 0, 0),
        "paired:colony:2016:partly lit" = c(1, 0, -1, -1, 0),
        "paired:colony:2016:fully lit" = c(1, 0, -1, 0, -1),
        "paired:colony or used:1980s:dark" = c(0, 1, 0, 0, 0),
        "paired:colony or used:2016:dark" = c(0, 1, -1, 0, 0),
        "paired:colony or used:2016:partly lit" = c(0, 1, -1, -1, 0),
        "paired:colony or used:2016:fully lit" = c(0, 1, -1, 0, -1)
      )
    )
  )
cumulative_prediction <- cumulative_logit %>%
  plogis() %>%
  apply(
    1, 
    function(x){
      c(
        Mean = mean(x), 
        LCL = quantile(x, 0.025) %>%
          unname(), 
        UCL = quantile(x, 0.975) %>%
          unname()
      )
    }
  ) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("Contrast") %>%
  extract(
    Contrast, 
    c("Dataset", "Status", "Survey", "Lights"), 
    "(.*):(.*):(.*):(.*)"
  ) %>%
  mutate(Lights = factor(Lights, levels = levels(status$Lights)))
```

```{r survey-prediction, fig.cap = "Estimated cumulative probability based on both surveys."}
ggplot(
  cumulative_prediction, 
  aes(
    x = Lights, 
    y = Mean, ymin = LCL, ymax = UCL, 
    colour = Survey, 
    shape = Dataset, linetype = Dataset
  )
) +
  geom_errorbar(position = position_dodge(width = 0.5)) +
  geom_point(position = position_dodge(width = 0.5)) +
  scale_x_discrete("Light status") +
  scale_y_continuous(
    "Estimated cumulative probability",
    labels = percent, 
    limits = 0:1
  ) +
  scale_color_manual(
    values = c("2016" = "black", "1980s" = "darkgrey")
  ) +
  facet_wrap(~Status)
```

## Effect of renovations

The renovation status is only available for churches from the 1980s survey, hence we can only use the paired dataset for this analysis. We added the renovation status as a fixed effect to the model. The interaction between lights status and renovation status would be relevant, but this gave a singular model. 

```{r renovation-ordinal}
renovation_model <- clmm(
  Status ~ Survey + Lights + Renovated + (1|Church), 
  data = status,
  subset = n == 2
)
renovation_lrt <- drop1(renovation_model, test = "Chisq")
renovation_lrt <- sprintf(
  "$\\chi^2_%i$=%.4g, p=%.2g", 
  renovation_lrt[-1, "Df"], 
  renovation_lrt[-1, "LRT"], 
  renovation_lrt[-1, "Pr(>Chi)"]
)
```

```{r renovation-coef}
renovation_or <- confint(renovation_model) %>%
  cbind(mean = coef(renovation_model)) %>%
  exp() %>%
  as.data.frame() %>%
  select(mean = 3, lcl = 1, ucl = 2) %>%
  mutate(
    print = sprintf("$OR:%.2f (%.2f; %.2f)$", mean, lcl, ucl)
  )
```

```{r renovation-contrast}
cumulative_logit <- contrasts(
  renovation_model, 
  contrast =  rbind(
    "contrast:dark-renovated:fully lit-unrenovated" = c(0, 0, 0, 0, 1, -1),
    "contrast:dark-renovated:partly lit-unrenovated" = c(0, 0, 0, 1, 0, -1),
    "colony:1980s:dark:no" = c(1, 0, 0, 0, 0, 0),
    "colony or used:1980s:dark:no" = c(0, 1, 0, 0, 0, 0),
    "colony:2016:dark:no" = c(1, 0, -1, 0, 0, 0),
    "colony:2016:partly lit:no" = c(1, 0, -1, -1, 0, 0),
    "colony:2016:fully lit:no" = c(1, 0, -1, 0, -1, 0),
    "colony or used:2016:dark:no" = c(0, 1, -1, 0, 0, 0),
    "colony or used:2016:partly lit:no" = c(0, 1, -1, -1, 0, 0),
    "colony or used:2016:fully lit:no" = c(0, 1, -1, 0, -1, 0),
    "colony:2016:dark:yes" = c(1, 0, -1, 0, 0, -1),
    "colony:2016:partly lit:yes" = c(1, 0, -1, -1, 0, -1),
    "colony:2016:fully lit:yes" = c(1, 0, -1, 0, -1, -1),
    "colony or used:2016:dark:yes" = c(0, 1, -1, 0, 0, -1),
    "colony or used:2016:partly lit:yes" = c(0, 1, -1, -1, 0, -1),
    "colony or used:2016:fully lit:yes" = c(0, 1, -1, 0, -1, -1)
  )
)
renovation_contrast <- cumulative_logit %>%
  "["(1:2, ) %>%
  exp() %>%
  apply(
    1, 
    function(x){
      c(
        mean = mean(x), 
        lcl = quantile(x, 0.025) %>%
          unname(), 
        ucl = quantile(x, 0.975) %>%
          unname()
      )
    }
  ) %>%
  t() %>%
  as.data.frame() %>%
  mutate(
    print = sprintf("$OR:%.2f (%.2f; %.2f)$", mean, lcl, ucl)
  ) %>%
  rownames_to_column("Contrast")
cumulative_prediction <- cumulative_logit %>%
  "["(-1:-2, ) %>%
  plogis() %>%
  apply(
    1, 
    function(x){
      c(
        Mean = mean(x), 
        LCL = quantile(x, 0.025) %>%
          unname(), 
        UCL = quantile(x, 0.975) %>%
          unname()
      )
    }
  ) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("Contrast") %>%
  extract(
    Contrast, 
    c("Status", "Survey", "Lights", "Renovation"), 
    "(.*):(.*):(.*):(.*)"
  ) %>%
  mutate(
    Lights = factor(Lights, levels = levels(status$Lights))
  )
```

### Interpretation

We find no evidence for an effect between surveys `r renovation_or$print[3]`, $LTR:$ `r renovation_lrt[1]`. The light status has a significant effect $LTR:$ `r renovation_lrt[2]`. We find an unclear effect of partly lit vs dark `r renovation_or$print[4]`, but the odds for no bats between fully lit and dark increases `r renovation_or$print[5]`. Addionally, after renovation, the odds for no bats increase `r renovation_or$print[6]`, $LTR:$ `r renovation_lrt[3]`. The predictions are given in fig. \@ref(fig:renovation-prediction). We could wonder which is worse: a fully lit, non renovated church or a dark, renovated church. After calculting the odds ratio `r renovation_contrast$print[1]`, we must conclude that we cannot make a hard statement that one is worse than the other.

```{r renovation-prediction, fig.cap = "Estimated cumulative probability for the renovation model"}
ggplot(
  cumulative_prediction, 
  aes(
    x = Lights, 
    y = Mean, ymin = LCL, ymax = UCL, 
    colour = Survey,
    linetype = Renovation
  )
) +
  geom_errorbar(position = position_dodge(width = 0.5)) +
  geom_point(position = position_dodge(width = 0.5)) +
  scale_x_discrete("Light status") +
  scale_y_continuous(
    "Estimated cumulative probability",
    labels = percent, 
    limits = 0:1
  ) + 
  scale_color_manual(
    values = c("2016" = "black", "1980s" = "darkgrey")
  ) +
  facet_wrap(~Status)
```

# Discussion

We demonstrate how cumulative logit (mixed) models[@ordinal] are useful to analyse ordinal data and give more insight in the data.
